### 2018-03-02 paper/sexSelectionSpacing-ver1.tex
Sample sizes were hardcoded into paper
Editing

% Fertility, Birth Spacing and the Determinants of sex-selective Abortions
% Understanding Sex Selection: Fertility, Abortions and Birth Spacing
% The Determinants of sex-selective Abortions: A New Method/Approach
% Sex Selection: A New Method for Estimating Determinants
% Fertility, Birth Spacing and Abortions in India
% Fertility, Birth Spacing and Sex Selection
% Fertility, Birth Spacing and sex-selective Abortions
% Fertility, Spacing and Sex Selection
% Fertility, Sex Selection, and Spacing
% Fertility, Sex Selection, and Birth Spacing
% Fertility, Birth Spacing and Missing Girls
% Where have all the girls gone: An analysis of sex-selective abortions

\title{Son Preference, Birth Spacing, and Sex-Selective Abortions%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support from the University of Washington Royalty Research Fund and the 
Development Research Group of the World Bank is gratefully acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
% Prior versions of this paper were presented under the title ``The 
% Determinants of sex-selective Abortions.''
}
}

\author{}

@@

\date{December 2017\\
\bigskip
Preliminary}

@@

Spacing between births has long been used as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for 
parents who wanted a son---but did not yet have one---was to have the next 
birth sooner.
Son preference was therefore often associated with shorter spacing after the 
births of girls than boys, which, in turn, was associated with worse 
health outcomes for girls.%
\footnote{
See, for example, 
\citet{Das1987}, \citet{Rahman1993}, \citet{Pong1994}, \citet{Haughton1996},
\citet{Arnold1997}, and \citet{Soest2012} on shorter birth spacing after
girls
and 
\citet{arnold98}, \citet{Whitworth2002}, \citet{Rutstein2005},
and \citet{Conde-Agudelo2006} on the association between shorter birth
spacing and worse health outcomes for girls.
Parents were also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{ben-porath76b,Das1987,Arnold1997,clark00,filmer09,Altindag2016}
}

What has not previously been appreciated is that the introduction 
of prenatal sex determination fundamentally changed the relationship 
between son preference and birth spacing.
Couples, who used to have short birth spacing because of son preference,
now have access to prenatal sex determination and sex-selective abortions,
but each sex-selective abortion they have substantially increases the space
between births. 
Birth spacing increases by approximately a year per abortion.
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of a spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which I set to 6 
months \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.%
\footnote{
The waiting time to conception varies by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}

As a result, we now have a situation where families with the \emph{strongest} son 
preference may show the \emph{longest} spacing after the birth of a daughter,
because they use sex-selective abortions.
But, to further complicate matters, we may still observe short spacing 
after the births of daughters as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

Spacing, by itself, can therefore no longer be used to measure son preference.
However, understanding birth spacing remains an important undertaking.
First, spacing is still useful in understanding son preference, 
if combined with the likelihood of observing a boy or a girl.
% Short spacing after a girl is indicative of son preference if observed
% births are close to the natural sex ratio, whereas long spacing is indicative
% of son preference if observed births are male biased.
Second, if spacing affects health outcomes for mother and/or children, it is 
important to understand what drives changes in spacing.
The better health outcomes for girls in the presence of sex selection may,
for example, be an unintended side-effect of the longer spacing that arise
from sex-selective abortions, rather than, as often assumed, be driven by a 
higher value placed on girls because of their lower number \citep{Lin2014,Hu2015}.
Third, spacing may be an important factor in parents' decisions, either
for preference or economic reasons.
This is especially important here, because even parents with strong
son preference may reverse their decision to use prenatal sex determination, 
and carry the next pregnancy to term whether male or female, as the duration 
from the prior birth becomes sufficiently sufficiently long.
Finally, we know less about what determines spacing behavior in 
developing countries than in developed countries.
With increasing numbers of women entering the labor force in developing
countries, understanding how timing decisions are made will be important
for the design of suitable policies \citep{Portner2018}.

In this paper, I introduce and apply a novel empirical method that directly incorporates 
the effects of sex-selective abortions on the likelihood of a son being born 
\emph{and} the duration between births.
The method can be used to analyse both situations with and without prenatal
sex selection.
Furthermore, my proposed method allow for the duration since the prior 
birth to affect the decision on sex selection.
By examining whether sex selection decisions change with spacing, we can 
draw a more nuanced picture of the degree of son preference.

% [Why India?]
I apply the method to birth histories for Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2006. 
India is a particularly interesting case.
On one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded.%
\footnote{
See \citet{das_gupta97}, \citet{Sudha1999},
\citet{Arnold2002}, \cite{retherford03b} and \citet{jha06}.
India is clearly not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, recent research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.%
\footnote{
This measure of son preference is commonly used in the literature. 
See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
}
In addition, there has been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971, and still is, but the 
first reports of prenatal sex determination did
not appear until 1982--83 \citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection 
became widespread after a senior government official's wife aborted a 
fetus that turned out to be male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
The Act is described in detail at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach 5 convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Maharashtra was the first state to pass a similar law in 1988.
}
Hence, the data make it possible to show how spacing between births and
sex ratios have changed with the introduction of prenatal sex determination,
and whether the ban affected the relationship between birth spacing and sex 
ratios.

@@

I divide women's reproductive lives into spells, which are the periods from 
one birth, or marriage for the first spell, to the following birth
(I define how spell duration is measured in Section \ref{sec:data}).
A basic problem is that some spells' duration are censored, either
because the survey took place before the birth that would end the spell 
or because the couple has finished childbearing.
The standard approach, which I follow here, is to use a hazard models to 
address the censoring of spells.

Prior research on birth spacing have used hazard models where the only 
exit state was the birth of a child. 
This works when there is no sex selection---making the sex of the child 
a random event---but not here because at least some couples use 
sex selection. 
To capture that the sex of children are no longer necessarily a random event, 
I instead use a competing risk framework with two possible exit states: 
either a boy or a girl is born.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Proportional hazard models, where covariates have an multiplicative 
effect on the hazard rate, are more efficient than non-proportional 
models, but only provided that the proportionality assumption holds.
Because proportional hazard models do not allow covariates to have 
different sized effects at different times within a spell,
they cannot capture differences in the shape of the hazard functions 
between different groups, and if the proportionality assumption does 
not hold the estimates may be biased.
It is unlikely---even in the absence of prenatal sex determination---that 
the effect of, for example, the sex composition of prior births have 
the same effect throughout the entire spell.
This is especially the case for later spells where different
sex composition of prior births can lead to substantial difference 
both in the likelihood of progressing to the next birth and
how soon couples want their next child if they are going to have one.
Any bias from the proportionality assumption is likely 
exacerbated by the introduction of prenatal sex determination.
First, sex-selective abortions affect birth spacing and use
vary across groups.
Second, the use of sex selection may vary within a spell, which means 
that the effects of covariates vary within the spell as well.
Because of these concerns I use a non-proportional specification.
The use of a non-proportional specification also has the benefit 
that it mitigates any potential effects of unobserved heterogeneity 
when used in conjunction with a flexible baseline hazard \citep{Dolton1995}.
In summary, my proposed model is a discrete time, non-proportional, competing risk 
hazard model with two exit states: either a boy or a girl is born.

For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$ and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: birth of a boy, $j=1$, or birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is
\begin{equation}
 h_{ijt} = \Pr (T_i=t, J_i=j \mid T_i \geq t; \mathbf{Z}_{it},\mathbf{X}_{i} ),
\end{equation}
where $T_i$ is a discrete random variable that captures when woman $i$'s birth occurs.
To ease presentation the indicator for spell number is suppressed.
The explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, include 
individual, household, and community characteristics discussed below.

@@

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, if the data are transformed so the unit of analysis is a time period, in 
this case 3 months periods, rather than the individual woman, the model can be 
estimated using a standard multinomial logit model.%
\footnote{
A potentially issue is that the multinomial model assumes that alternative 
exit states are stochastically independent,
also known as the Independence of Irrelevant Alternatives (IIA) assumption.
This assumption rules out any individual-specific unmeasured or 
unobservable factors that affect both the hazard of having a girl and the 
hazard of having a boy.
To address this issue the estimations include a proxy for fecundity
discussed in Section \ref{sec:data}.
In addition, the multivariate probit model can be used as an alternative
to the multinomial logit because the IIA is not imposed \citep{han90}.
The results are essentially identical between these two models and
available upon request.
}
In the reorganized data the outcome variable is 0 if the
woman does not have a child in a given period (the base outcome), 1 if 
she gives birth to a son in that period, and 2 if she gives birth to 
a daughter in that period.

The main downside of this set-up is that direct interpretation of the estimated 
coefficients is challenging;
the coefficients show the change in hazards relative to the base outcome, 
here no birth, rather than simply the hazard of an event.
Hence, a positive coefficient does not necessarily imply that the associated 
exit state becomes likely as the explanatory variable increases because the 
probability of the other exit state may increase even more \citep{thomas96}.
It is, however, straightforward to calculate the predicted probabilities of 
having a boy and of having a girl for each $t$ within a spell, conditional on 
a set of explanatory variables and not having had a child before that period.
The predicted probability of having a boy in period $t$ for a given set of 
explanatory variable values, $\mathbf{Z}_k$ and $\mathbf{X}_k$, is
\begin{equation}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})},
\label{eq:probability_boy}
\end{equation}
and the predicted probability of having a girl is
\begin{equation}
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}.
\label{eq:probability_girl}
\end{equation}
Within each period the probability of not having a birth in period $t$
is $1-P(b_{t})-P(g_{t})$.

@@

An important issue is that the probability of ever having a next birth varies 
across groups.
For example, as the parity progression literature shows, couples with stronger 
son preference are less likely, for a given number of prior births, to ever have 
a next birth if they already have one or more sons.
Direct comparison of standard survival curves therefore tells us little 
about how the spread of sex selection affect birth spacing across, for
example, different sex compositions of prior children.
To overcome this problem I condition on parity progression when examining
birth spacing using the survival curves.
This allows me to calculate measures, such as median spell duration, for
the women with a given set of characteristics who will eventually give
birth in a spell.%
\footnote{
It is important to note that this is not the same as simply calculating 
the median \emph{observed} spell lengths among women who have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.
}
This number can then be compared across groups to understand how birth
spacing changes with the introduction of sex selection.
Obviously, this approach only works if the spell length covered is 
sufficiently long that only very few women would be likely to give
birth later.
In addition to calculating conditional spell length for different 
percentages, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

Finally, with the predicted probabilities of having a boy and of having 
a girl for each $t$ within a spell, is it easy to calculate the estimated 
percentage of children born that are boys, $\hat{Y}$, at each $t$  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting across all $t$ 
gives the predicted percent boys born over the entire spell.
In addition to the predicted percent boys over the entire spell, 
I also present graphs of how the percentage boys born vary across time
within a spell, together with the associated confidence interval for given 
values of explanatory variables calculated using the Delta method.
These are combined with standard survival curves which shows the ``weight''
that each period should be given, or, in other words, how many births take
place in each period.

@@

The data come from the four rounds of the National Family Health Survey 
(NFHS-1, NFHS-2, NFHS-3, and NFHS-4),
collected in 1992--93, 1998--99, 2005--2006, and 2015--2016.%
\footnote{
A delay in the survey for Tripura means that NFHS-2 has a small number of observation 
collected in 2000.
}
The surveys are large: NFHS-1 covered 89,777 ever-married women 
aged 13--49 from 88,562 households,
NFHS-2 covered 90,303 ever-married women aged 15--49 from 92,486 households,
NFHS-3 covered 124,385 never-married and ever-married women aged 
15--49 from 109,041 households,
and 
NFHS-4 covered 699,686 never-married and ever-married women aged
15--49 from 601,509 households surveyed.

I exclude visitors to the household, as well as
women married more than once, divorced, or not living with their husband,
women with inconsistent information on age of marriage,
and those with missing information on education.
Women interviewed in NFHS-3 or NFHS-4 who were never married or where gauna 
had not been performed were also dropped.
The same goes for women who had at least one multiple birth,
reported having a birth before age 12, had a birth before marriage, or
a duration between births less than 9 months.
Women who reported less than 9 months between marriage and first birth
remain in the sample---unless they are dropped for another reason---because 
10 to 20 percent fall into this category across the 4 surveys.
Although it is possible that some of these births are premature, the high number of
women who report a birth less than 6 months after marriage indicates that for
a majority conception likely occurred before marriage.

Finally, I restrict the sample to Hindus,
who constitute about 80 percent of India's population.
If use of sex selection differ between Hindu and other religions, such 
as Sikhs, assuming that the baseline hazard is the same would lead to bias.
The other groups are each so small relative to Hindus that it is not
possible to estimate different baseline hazards for each group.
Furthermore, the other groups are so different in terms of background and son 
preference that combining them into one group would not make sense.

There are four advantages to using the NFHS.
First, surveys enumerators pay careful attention to spacing between births and
probe for ``missed'' births.
Second, no other surveys cover as long a period in the same amount of detail.
The four NFHS rounds allow me to show the development in spacing and 
sex ratio from before sex-selective abortions were available until 2016.
Third, NFHS has birth histories for a large number of women.%
\footnote{
The Special Fertility and Mortality Survey appears to cover a much large number of households
than the three rounds of the NFHS combined, but \citet{jha06} only use the births that 
took place in 1997 making their sample sizes by parity smaller than here.
Their sample consists of 133,738 births of which 38,177 were first
born, 36,241 second born, and 23,756 were third born.
The differences in results for first born children are discussed in the online 
Appendix.
}
Finally, even if probing for missing births may not completely eliminate recall error,   
the overlap in cohorts covered and the large sample size make it possible to establish 
where recall error remains a problem.

>>>>

\title{Son Preference, Birth Spacing, and Sex-Selective Abortions%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support from the University of Washington Royalty Research Fund and the 
Development Research Group of the World Bank is gratefully acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

@@

\date{March 2018\\
\bigskip
Preliminary}

@@

Parents' spacing between births has long served as a measure of son preference 
\citep{Leung1988}.
Before prenatal sex determination became available, the only recourse for 
parents who wanted a son---but did not yet have one---was to have the next 
birth sooner.
Son preference is therefore often associated with shorter spacing after the 
births of girls than boys 
\citep{Das1987,Rahman1993,Pong1994,Haughton1996,Arnold1997,Soest2012}.
Shorter spacing is, in turn, associated with worse health outcomes for girls 
\citep{arnold98,Whitworth2002,Rutstein2005,Conde-Agudelo2006}.
\footnote{
Parents are also more likely to cease childbearing after the birth of 
a son than after a daughter 
\citep{ben-porath76b,Das1987,Arnold1997,clark00,filmer09,Altindag2016}.
}

The introduction of prenatal sex determination fundamentally changed the 
relationship between son preference and birth spacing.
Couples, who before had shorter birth spacing because of son preference,
now have access to prenatal sex determination and sex-selective abortions,
and each abortion increases birth spacing by approximately a year.
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which I set to six 
months \citep{Wang2003}.%
\footnote{
The waiting time to conception varies by woman, but even if it is, say, one 
month, the additional space between births would still be six months per abortion.
}
Finally, sex determination tests are reliable only from 3 months 
of gestation onwards.

As a result, we now have a situation where families with the 
\emph{strongest} son preference may have \emph{longer} spacing after 
the birth of a daughter,
because of their use of sex-selective abortions.
To further complicate matters, we may still observe short spacing 
after the births of daughters as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.
Also, even if they have access to prenatal sex determination, couples 
with stronger son preference may still try to conceive faster after daughters, 
and may even shorten the time until they try to conceive, knowing that 
they may have to go through multiple pregnancies and sex-selective
abortions before they conceive a son.

Spacing, by itself, can therefore no longer be used as a direct measure of 
son preference.
However, understanding birth spacing remains a critical undertaking.
First, birth spacing is still useful in understanding son preference, 
if combined with the likelihood of observing a boy or a girl.
Second, if spacing affects health outcomes for mother and children, it is 
essential to understand what drives changes in spacing.
The better health outcomes for girls in the presence of sex selection may,
for example, be an unintended side-effect of the longer spacing that arises
from sex-selective abortions, rather than, as often assumed, be driven by a 
higher value placed on girls because of their lower number \citep{Lin2014,Hu2015}.
Third, the duration between births may be an important factor in parents' 
decisions, either for preference or economic reasons.
Whether duration plays a role in parents' decisions is especially important 
here, because even parents with strong son preference may reverse their 
decision to use prenatal sex determination, and carry the next pregnancy 
to term whether male or female, as the duration from the prior birth 
becomes sufficiently long.
Finally, we know less about what determines spacing behavior in 
developing countries than in developed countries.
With increasing numbers of women entering the labor force in developing
countries, understanding how timing decisions are made will be necessary
for the design of suitable policies \citep{Portner2018}.

In this paper, I introduce and apply a novel empirical method that directly incorporates 
the effects of sex-selective abortions on the likelihood of a son being born 
\emph{and} the duration between births.
The method can be used to analyze both situations with and without prenatal
sex selection.
Furthermore, my proposed method allows for the time since the previous 
birth to affect the decision on sex selection.
By examining whether sex selection decisions change with spacing, we can 
draw a more nuanced picture of the degree of son preference.

% [Why India?]
I apply the method to birth histories for Hindu women, using data from 
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016. 
India is a particularly compelling case.
On the one hand, India has seen dramatic increases in the males-to-females ratio 
at birth over the last three decades as access to prenatal sex determination 
expanded 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06}.%
\footnote{
India is not alone; both China and South Korea saw
significant changes in the sex ratio at birth over the same period 
\citep{Yi1993,park95}.
}
On the other hand, research suggests that son preference in 
India, when measured as ideally having more boys than girls, is decreasing 
over time and with higher education \citep{bhat03,pande07}.
% %
% \footnote{
% This measure of son preference is commonly used in the literature. 
% See, for example, \citet{clark00}, \citet{Jensen2009}, and \cite{Hu2015}.
% }
There have also been substantial changes in access and legality of
prenatal sex determination in India over the period covered.
Abortion has been legal in India since 1971 and still is, but the 
first reports of prenatal sex determination did
not appear until 1982--83 \citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection 
became widespread after a senior government official's wife aborted a 
fetus that turned out to be male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques 
(PNDT) Act, making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics apparently operate with little risk of legal action 
\citep{Sudha1999}.
Maharashtra was the first state to pass a similar law in 1988.
}
Hence, the data make it possible to show how the spacing between births and
sex ratios have changed with the introduction of prenatal sex determination,
and whether the ban affected the relationship between birth spacing and sex 
ratios.

@@

The unit of analysis is individual spells, the period from one birth, or marriage for
the first spell, to the next.
For some spells, the duration is censored, either because the survey
took place before the birth that would end the spell or because the couple
has finished childbearing.
The standard approach to address the censoring of spells, which I follow here, is 
to use a hazard model.
To capture the effects of sex selection, I extend the standard hazard model 
to allow for multiple exit states and a non-proportional hazard specification.

First, I use a competing risk framework with two possible exit states, either a boy
or a girl is born, because access to prenatal sex determination means that
the sex of children born is no longer necessarily a random event.
Prior research on birth spacing has used hazard models where the only 
possible exit state was the birth of a child. 
Having only one exit state works when there is no sex selection---making the 
sex of the child a random event---but not here because at least some couples use 
sex selection.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, I use a non-proportional hazard specification because
sex composition and the use of sex selection are likely to affect 
the shape of the hazard functions across groups.
Proportional hazard models, where covariates have a multiplicative 
effect on the hazard rate, are more efficient than non-proportional 
models, but only provided that the proportionality assumption hold.
If the proportionality assumption does not hold, the estimates
are biased.
It is unlikely---even in the absence of prenatal sex determination---that 
the effect of, for example, the sex composition of previous births have 
the same effect throughout the entire spell.
Assuming that the effect of sex composition is the same throughout a spell
is especially problematic for higher-order spells where different
sex composition of previous births can lead to substantial difference 
both in the likelihood of progressing to the next birth and
how soon couples want their next child if they are going to have one.
The introduction of prenatal sex determination is likely to exacerbate 
any bias from the proportionality assumption.
First, sex-selective abortions affect birth spacing, and use
varies across groups.
Second, the use of sex selection may vary within a spell, which means 
that the effects of covariates vary within the spell as well.
The use of a non-proportional specification also mitigates any potential 
effects of unobserved heterogeneity when used in conjunction with a 
flexible baseline hazard \citep{Dolton1995}.

In summary, my proposed model is a discrete time, non-proportional, competing risk 
hazard model with two exit states: either a boy or a girl is born.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$ and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: birth of a boy, $j=1$, or birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is
\begin{equation}
 h_{ijt} = \Pr (T_i=t, J_i=j \mid T_i \geq t; \mathbf{Z}_{it},\mathbf{X}_{i} ),
\end{equation}
where $T_i$ is a discrete random variable that captures when woman $i$'s birth occurs.
To ease presentation the indicator for spell number is suppressed.
The explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, include 
individual, household, and community characteristics discussed below.

@@

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, transforming the data, so each observation is a period---here equal
to three months---the model can be estimated using a standard multinomial logit model.
if the data are transformed so the unit of analysis is a time period, in 
this case 3 months periods, rather than the individual woman, the model can be 
estimated using a standard multinomial logit model.%
\footnote{
A potentially issue is that the multinomial model assumes that alternative 
exit states are stochastically independent,
also known as the Independence of Irrelevant Alternatives (IIA) assumption.
This assumption rules out any individual-specific unmeasured or 
unobservable factors that affect both the hazard of having a girl and the 
hazard of having a boy.
To address this issue the estimations include a proxy for fecundity
discussed in Section \ref{sec:data}.
In addition, the multivariate probit model can be used as an alternative
to the multinomial logit because the IIA is not imposed \citep{han90}.
The results are essentially identical between these two models and
available upon request.
}
In the reorganized data the outcome variable is 0 if the
woman does not have a child in a given period (the base outcome), 1 if 
she gives birth to a son in that period, and 2 if she gives birth to 
a daughter in that period.

The main downside of this set-up is that interpretation of the estimated 
coefficients is challenging;
the coefficients show the change in hazards relative to the base outcome, 
here no birth, rather than the hazard of an event.
Hence, a positive coefficient does not necessarily imply that the associated 
exit state becomes more likely as the explanatory variable increases because the 
probability of the other exit state may increase even more \citep{thomas96}.
It is, however, straightforward to calculate the predicted probabilities of 
having a boy and of having a girl for each $t$ within a spell, conditional on 
a set of explanatory variables and not having had a child before that period.
The predicted probability of having a boy in period $t$ for a given set of 
explanatory variable values, $\mathbf{Z}_k$ and $\mathbf{X}_k$, is
\begin{equation}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})},
\label{eq:probability_boy}
\end{equation}
and the predicted probability of having a girl is
\begin{equation}
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}.
\label{eq:probability_girl}
\end{equation}
Within each period the probability of not having a birth in period $t$
is $1-P(b_{t})-P(g_{t})$.

@@

An important issue is that the probability of ever having a next birth varies 
across groups.
For example, as the parity progression literature shows, couples with stronger 
son preference are less likely, for a given number of prior births, to ever have 
a next birth if they already have one or more sons.
Direct comparison of standard survival curves, therefore, tells us little 
about how the spread of sex selection affects birth spacing across, for
example, different sex compositions of prior children.

To overcome this problem, I condition on the likelihood of parity progression 
when examining birth spacing.
If, for example, women with a given set of characteristics have an 80\%
chance of ending a spell with a birth their median spell duration is when
their survival curve shows that 40\% are predicted to have exited the spell.%
\footnote{
It is important to note that this is not the same as simply calculating 
the median \emph{observed} spell lengths among women who have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.
}
Median durations can then be averaged across women with different 
characteristics using the probabilities of ending a spell with a child
as weights.
How well this approach works depends on whether the spell length covered is 
sufficiently long that only very few women would be likely to give
birth later.
I discuss the choice of spell length below.
In addition to calculating conditional spell length for different 
percentages, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

Finally, with the predicted probabilities of having a boy and of having 
a girl for each $t$ within a spell, is it easy to calculate the estimated 
percentage of children born that are boys, $\hat{Y}$, at each $t$  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting across all $t$ 
gives the predicted percent boys born over the entire spell.
In addition to the predicted percent boys over the entire spell, 
I also present graphs of how the percentage boys born vary across time
within a spell, together with the associated confidence interval for given 
values of explanatory variables calculated using the Delta method.
Finally, I include standard survival curves, which shows the ``weight''
that each period has, or, in other words, how many births take place in
each period.

@@

The data come from the four rounds of the National Family Health Survey 
(NFHS-1, NFHS-2, NFHS-3, and NFHS-4),
collected in 1992--93, 1998--99, 2005--2006, and 2015--2016.%
\footnote{
A delay in the survey for Tripura means that NFHS-2 has some observation 
collected in 2000.
}
The surveys are large: NFHS-1 covered 89,777 ever-married women 
aged 13--49 from 88,562 households;
NFHS-2 covered 90,303 ever-married women aged 15--49 from 92,486 households;
NFHS-3 covered 124,385 never-married and ever-married women aged 
15--49 from 109,041 households;
and 
NFHS-4 covered 699,686 never-married and ever-married women aged
15--49 from 601,509 households surveyed.

I exclude visitors to the household, as well as
women who have been married more than once, divorced, or who are 
not living with their husband,
women with inconsistent age at marriage,
and those with missing information on education.
Women interviewed in NFHS-3 or NFHS-4 who were never married or where gauna 
had not been performed were also dropped.
The same goes for women who had at least one multiple births,
reported having a birth before age 12, had a birth before marriage, or
duration between births of less than nine months.
Women who reported less than nine months between marriage and first birth
remain in the sample---unless they are dropped for another reason---because 
10 to 20 percent fall into this category across the four surveys.
Although it is possible that some of these births are premature, the high number of
women who report a birth less than six months after their marriage indicates that for
a majority conception likely occurred before marriage.

Finally, I restrict the sample to Hindus,
who constitute about 80 percent of India's population.
If the use of sex selection differ between Hindu and other religions, such 
as Sikhs, assuming that the baseline hazard is the same would lead to bias.
The other groups are each so small relative to Hindus that it is not
possible to estimate different baseline hazards for each group.
Furthermore, the other groups are so different in background and son 
preference that combining them into one group would not make sense.

There are four advantages to using the NFHS.
First, surveys enumerators pay careful attention to spacing between births and
probe for ``missed'' births.
Second, no other surveys cover as extended a period in the same amount of detail.
The four NFHS rounds allow me to show how spacing and 
sex ratio changed from before sex-selective abortions were available until 2016.
Third, NFHS has birth histories for a large number of women.
% %
% \footnote{
% The Special Fertility and Mortality Survey appears to cover a much large number of households
% than the three rounds of the NFHS combined, but \citet{jha06} only use the births that 
% took place in 1997 making their sample sizes by parity smaller than here.
% Their sample consists of 133,738 births of which 38,177 were first-born, 
% 36,241 second-born, and 23,756 were third-born.
% }
Finally, even if probing for missing births does not eliminate recall error,   
the overlap in cohorts covered and the large sample size makes it possible 
to establish where recall error remains a problem.

Recall error arises mainly from child mortality when respondents are 
reluctant to discuss deceased children.
Systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, is especially problematic because 
it biases both spacing and sex ratios.
Probing catches many missed births, but systematic recall error is 
still a potentially substantial problem.
